Honey bees vary collective decision making in different landscapes

### 2022-08-04 inst/publication/Manuscript/paper.Rmd
updated MS with Elli edits

---
title: "Honey bees vary collective decision making in different landscapes"
# author: "Joseph Palmer, Ash Samuelson, Rich Gill, Elli Leadbeater and Vincent A. A. Jansen"
output:
  word_document
always_allow_html: true
bibliography: reffs.bib
csl: science.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(showtext)
showtext_auto()

@@

Collective behaviours emerge from the interactions between individuals which vary in their use of different sources of information. In social animals, collective behaviours emerge from individuals following behavioural rules which produce a collective level response, yet some conditions make it more profitable for individuals to abandon such structures and use individual rather than collective information. In large animal groups, where the ability to track and analyse the interactions between individuals is limited, our understanding of how decisions influence collective processes is largely theoretical. Empirically evaluating how individual decisions drive group level behaviours is an important objective with wider consequences for our understanding of the evolution of such complex collective behaviours. Here, we explore information use in a model social insect, the honey bee (*Apis* Spp.) and show empirically how variation in individual honey bee's decision making influences collective decision making at a colony level. By developing a method to assess a colonies use of the famous waggle dance, we quantify to what extent colonies use the dance to forage collectively and find collective foraging using the dance varies widely over different landscapes. These results demonstrate that collective decision making varies in response to environmental conditions. Our methods provide new insights into how individual variation in information use shapes an emergent collective process and provides opportunities to further probe the environmental drivers of collective foraging.

@@

In group living animals, many decisions are taken collectively by integrating information from individuals to produce a social behaviour which extends beyond that of the individual units [@Sumpter2006]. Such systems are self-organised, relying on the constituents using socially provided information and following behavioural 'rules' to produce a collective outcome [@Couzin2003;@Bonabeau1997]. However, in addition to information provided by others, individuals also have knowledge of their own (private information) and so can choose to work collectively or individually [@Dall2005;@Rieucau2011;@Gruter2014]. The benefits of working collectively versus independently vary across environmental contexts and individuals often adjust their own use of information accordingly [@Dechaume-Moncharmont2005;@Beekman2008]. The impact these individual variations in information use have at a group level, however, is less well understood and difficult to empirically evaluate [@Rieucau2011;@Giardina2008;@Delgado2018].

Understanding the collective ramifications of individual variation in information use is not only important for deciphering how social animals collectively regulate information use [@Gruter2014], relating this to the environmental context of the behaviour is key to illuminating the ultimate evolutionary drivers of collective processes [@Price2015]. However, establishing the group level effects of individual variation in information use is not trivial and requires measuring information use across animal groups; an often laborious process which requires monitoring individuals' behaviours and interactions. This becomes logistically impossible in social insects, which display some of the most intricate group decision-making mechanisms and consist of often thousands of individuals, each using a mix of individual and collective information to make decisions [@Bonabeau1997].

Social insects are defined by their highly coordinated collective responses and so provide ideal systems to investigate the impact of individual variation in information use on emergent collective processes. Nevertheless, the sheer quantity of individuals in many colonies is a barrier to further understanding these dynamics. Honey bees (Apis spp.) provide an opportunity to sidestep this problem of scale as they provide a detailed record of colony foraging activity in the form of their unique waggle dance -unparalleled in even a single other social insect [@Seeley1995]. Using the dance, individuals share information about profitable resource locations. A series of simple rules which determine when and how much bees perform waggle dances (Fig. 1) mean that choices between feeding sites occur at the level of the group rather than the individual [@Seeley1991;@Seeley1994A;@Hasenjager2022]. Despite this, a wealth of research shows that individual foragers frequently do not use dance information to find forage sites and instead rely on their own knowledge [@Gruter2009]. Moreover, studies investigating the value of the dance have shown that when prevented from communicating via dances, honey bee colonies can achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002;@Dornhaus2004;@Gruter2013;@Price2019].

```{r, fig.cap="Figure 1. **The honeybee waggle dance carries information about the location of a resource.** The duration of the waggle run indicates the distance to the resource and the angle of the dance relative to the vertical indicates the direction of the resource (circle in right panel), relative to the direction of the sun [@Frisch1993]. Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Overall resource “quality” -the net energetic gain of a foraging trip- is provided through the number of waggle runs performed [@Frisch1993;@Boch1956;@Esch1961;@Seeley1994;@Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Seeley1991], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A]."}
knitr::include_graphics("../results/figures/manuscript_figures/fig1.png")
```

The waggle dance provides a model system to study the group effects of collective and individual information use. Although it is likely that individual variation in dance use translates into variations in collective decision making at the colony level [@Donaldson-Matasci2012;@Price2019], this has not been empirically established. Individual based models, which explore the collective effects of individuals following different information streams provide an insight into how individual behaviours scale to the level of the group [@Dornhaus2006;@Beekman2008], however, these are difficult to relate to empirical observations and so are limited to showing what a colony under certain conditions might do, rather than quantifying what they are doing. Empirically determining how dance information is used at a colony level requires new methods which can identify the behaviours of individual foragers and integrate them with the mechanisms by which the dance functions to make collective decisions. Previously, this has only been quantifiable by documenting foragers' individual search history, which requires tracking individual bees and is time -and labour- intensive [@Seeley1995].

Here, we investigate how variation in individual information use influences the collective foraging behaviour of honey bees. Using a model based on well-established foraging behaviours [@Seeley1986;@Seeley1991;@Seeley1992;@Seeley1994;@Seeley1994A;@Seeley1995], we examine the distribution of waggle run durations and infer from its shape the relative contribution of dances for resources identified by individual search - “scouting”- versus dances for resources found by following dances - “recruitment”- [@Seeley1995]. This allows us to quantify the extent to which colonies use dance communication and thus forage collectively. To investigate how information use varies across environment, we fit our model to waggle dance observations from 20 sites in an agri-rural and urban environment (10 in each) and evaluated the landscape characteristics which best explained the variance in dance use.

@@

```{r, fig.cap="Figure 2. **Simulating honey bee foraging**. In our simulation model with scouting only (A), foragers leave the hive on a search path (white lines) and continue until they encounter a resource (circles, colours indicate different resource quality). When foraging with recruitment (B) foragers continue to identify resources in scouting trips (white lines) and convey this information on the dance floor (brown disc) where foragers can sample dances reporting on scouting and recruiting trips and follow these directions (yellow lines). (C) Complementary cumulative frequencies of foraging distances reported from scouting and recruit trips. Note the difference in the shape of the distributions. The scout distribution is best fit by an exponential (grey fit line), the recruit distribution by a Rayleigh distribution (black line)"}
knitr::include_graphics("../results/figures/manuscript_figures/fig2.png")
```

@@

```{r, fig.cap="Figure 3. **The rationale of the foraging model**. The distances of resources encountered by scouts are distributed exponentially (A). These dances are advertised on the dance floor (B). Dances for resources that are closer or higher in quality are repeated more often (C). As a consequence, dances for more profitable resource are over-represented and sampling foragers are biased to the more profitable resources (D). After successfully visiting advertised resources, recruits also dance for them leading to further amplification of this bias towards the most profitable resource in the vicinity of the hive (E). The distances of recruiting trips are than distributed through a Rayleigh distribution (F). Recruits also report the locations on the dance floor (B) and repeat their runs more often depending on the profitability of the location (C), leading to a distribution of durations of recruit dances (G). By taking together the dance distributions for scouts (D) and recruits (G) the distributions of all dances on the dance-floor can be found. The distances reported on the dance floor this are a mixture of the scout and recruiting trips and can be calculated from the distance distributions of the scouting and recruiting trips, taking the reporting bias into account (see Materials and Methods for detail)."}
knitr::include_graphics("../results/figures/manuscript_figures/fig3.png")
```

To evaluate the use of waggle dance information and individual search in honeybee colonies foraging in 'natural' landscapes, we analysed a pre-existing dataset of 2827 waggle dance observations from 20 observation hives, recorded between April-September 2017, (previously described in [@Samuelson2021]). Hives that contributed to this dataset had been situated at different locations in South East England (see Materials and Methods, figure 4A) and visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data was recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually [@Couvillon2012;@Schurch2019;@Samuelson2021] to extract waggle run durations. Using this data, for each site we fitted both the collective and individual models and used model selection to determine which provided the better explanation of the data, and (if the collective model provided a better fit) to quantify the relative use of social information through estimating the parameter $p$. In each case, we calculated the goodness-of-fit using a Kolmorgorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data [@Goldstein2004;@Clauset2009].

@@

```{r, fig.cap="Figure 4. **The honey bee foraging model fitted to data from 20 hives**. (A) Location of study hives in Southern England, shaded area in the main plot indicates Greater London. For 16 hives for the collective foraging model provided best explanation (black circles) or for 4 hives the individual search model provided the best explanation (red triangles) as indicated by lowest AIC score. (B). Distribution of goodness of fit confidence values for each model fit to waggle run durations from each site. The p-value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data, the red dashed line marks the significance threshold of 0.05. For values exceeding the threshold there is no statistically significant difference between the model and the data, indicating the model provides a good fit. For the hive in (C) the individual model (blue line) provided a better fit than the collective foraging model (red line). For the hive in (D) the collective foraging model (red line) provided a better fit than the individual model (blue line). The typical “hump” in the distribution in (D) which is indicative of contribution of recruitment dances (compare to Fig 3H). Panels show the compliment cumulative frequencies with binned frequency distributions as inset."}
knitr::include_graphics("../results/figures/manuscript_figures/fig4.png")
```

@@

Further quantification of the use of waggle-dance recruitment within all colonies, as a proportion of all foraging trips, can be achieved by extracting the estimated proportion of recruit trips, $1-p$, for each site. Since our sites varied in land-use characteristics and potentially thus forage distributions (although in this case, not by design for this study), we investigated whether these estimates might correlate with land-use. We first classified the different land-use types of the area surrounding each site [@Samuelson2018] to obtain a standardised land-use profile for the urban and agri-rural environments separately as many land-use types present in urban areas do not occur in agri-rural environments and vice versa (see Materials and Methods). We then performed a Partial Least Squares (PLS) analysis [@Carrascal2009] (see Materials and Methods) to determine the principal components that represent combinations of land-use types which explained the most variation in waggle dance use within agri-rural and urban environments. As for one of our sites neither model provided a plausible description, this site was removed from the PLS analysis.

@@

```{r, fig.cap="Figure 5. **Collective foraging correlates with land-use**. Estimated waggle dance use for each site against the first principal component derived from a Partial Least Squares analysis of land-use type. Beta regression shows the relationship (black line) between first principal component and waggle dance use, with 95% CI shown by the grey shaded area for agri-rural (A) and urban landscapes (C). The correlations between first principal component and each land-use type are shown for agri-rural (B) and urban landscapes (D). Correlations outside the shaded area significantly contribute to the first principal component. Colours correspond to the land use as shown in maps (circular insets) for selected sites. NAUMGS (resp. NAMGS) stands for non-agricultural unmanaged (resp. managed) green space"}
knitr::include_graphics("../results/figures/manuscript_figures/fig5.png")
```

@@

load("../results/area_coverage.RData")

@@

Goodness of fit is assessed using the two-sample Kolmorgorov-Smirnov (KS) test [@Goldstein2004] and implemented in R using the ks.boot function of the package 'Matching' in R [@Sekhon2011].

@@

All data and code is available on request and will be made available on zenodo prior to publication.

>>>>

---
title: "Honey bees vary collective decision making in different landscapes"
# author: "Joseph Palmer, Ash Samuelson, Rich Gill, Elli Leadbeater and Vincent A. A. Jansen"
output:
  word_document
always_allow_html: true
bibliography: reffs.bib
csl: science.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
devtools::load_all()
library(showtext)
showtext_auto()

@@

Honeybee foraging is an extraordinary collective behaviour which emerges from one of the most sophisticated communication systems in the animal kingdom: the waggle dance. Through a series of feedback loops, this elaborate behaviour -unparalleled in even a single other social insect- allows colonies to decide where to forage without requiring any individual to directly compare resources. This behaviour is a textbook example of an emergent collective process, yet a growing body of evidence now shows that using the dance does not always improve foraging success and sometimes not having access to the dance even increases it. Although the value of dancing varies across environments, it is not known if and to what extent colonies can adjust dance use in response. Here, develop a method to quantify dance use and show that colonies vary the use of the waggle dance in different landscapes, with some colonies relying on dance use almost exclusively and others not at all. Further probing of the drivers of this variation in dance use found a colonies use of the dance correlates with different landscape characteristics, suggesting colonies adjust collective decision-making in response to environmental conditions. Our results provide new evidence of flexible dance use in honey bee colonies and our methodology outlines a means to identify the ecological conditions in which honeybee colonies rely on dance communication, opening the door to large-scale experimental exploration of the selection pressures that may have driven the evolution of this remarkable collective behaviour.

@@

In group living animals, many decisions are taken collectively by integrating information from multiple individuals to produce a social behaviour which extends beyond that of the individual units [@Sumpter2006]. Such systems are self-organised, relying on the constituents using socially provided information and following behavioural 'rules' to produce an emergent collective outcome [@Couzin2003;@Bonabeau1997]. The behavioural architectures which produce these emergent collective behaviours have in some cases, such as in social insects, grown to dazzling levels of complexity. However, this complexity is not always beneficial. The benefits of working collaboratively versus independently vary across environmental contexts and individuals often adjust their own use of information accordingly [@Dechaume-Moncharmont2005;@Beekman2008]. Although the benefits of collective behaviours are known to vary across ecological conditions, in many cases we have a more limited understanding of if and how a group can flexibly use collective information to make use of the most profitable strategy under different conditions.

The extraordinary waggle dance communication system of the honey bee (*Apis* spp.) is a key example of a behavioural mechanism which produces emergent collective decisions [@Seeley1995]. Within colonies, a series of simple rules which determine when and how much bees perform dances (Fig. 1) mean that choices between feeding sites are made by the group rather than individuals [@Seeley1991;@Seeley1994A;@Hasenjager2022]. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the richer of two equidistant sources [@Seeley1991], or the closer of two equally rich sources [@Hasenjager2022]. This mechanism allows a colony to pool individual assessments of the landscape and processes them collectively to identify the most profitable resources to focus on, all without the need for any individual to compare resources.

```{r, fig.cap="Figure 1. **The honeybee waggle dance carries information about the location of a resource.** The duration of the waggle run indicates the distance to the resource and the angle of the dance relative to the vertical indicates the direction of the resource (circle in right panel), relative to the direction of the sun [@Frisch1993]. Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Overall resource “quality” -the net energetic gain of a foraging trip- is provided through the number of waggle runs performed [@Frisch1993;@Boch1956;@Esch1961;@Seeley1994;@Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Seeley1991], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A]."}
knitr::include_graphics("../Results/figures/manuscript_figures/fig1.png")
```

Despite it's complexity, research has shown that the ***value*** the waggle dance provides to foraging success varies according to environmental context and colonies without access to dance information achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002;@Dornhaus2004;@Gruter2013;@Price2019]. Interestingly, studies have shown that individuals often ignore the dance when looking for forage sites [@Gruter2009]. Whilst it likely that individual variation in dance use translates into variations in collective decision making at the colony level [@Donaldson-Matasci2012;@Price2019], this has yet to be empirically established. Identifying if and how colonies vary waggle dance ***use*** is not only important for deciphering how honey bees collectively respond to changes in environmental conditions [@Gruter2014], relating this to the environmental context of the behaviour is key to illuminating its ultimate evolutionary drivers [@Price2015].

Quantifying dance use within colonies requires measuring dance use across the foraging workforce [@Seeley1995]. This has only previously been quantifiable by documenting foragers' individual search history, which requires tracking individual bees and is time -and labour- intensive [@Seeley1995]. Individual based models, which explore the collective effects of individuals following different behaviours, provide an insight into how individual information use scales to the level of the group [@Dornhaus2006;@Beekman2008], however, these are difficult to relate to empirical observations and so are limited to showing what a colony under certain conditions might do, rather than quantifying what they are doing. Empirically determining how dance information is used at a colony level requires new methods which can identify the behaviours of individual foragers and integrate them with the mechanisms by which the dance functions to make collective decisions.

Here, we derive a mathematical model to quantify waggle dance use in colony foraging by identifying the proportion of foraging trips that are made by bees recruited through waggle dances. To validate our model, we fit it to waggle dance observations collected from 20 sites across and agri-rural and urban environment (10 in each). We then use our model to explore how variations in landscape structure influences dance use in colony foraging.

@@

```{r, fig.cap="Figure 2. **Simulating honey bee foraging**. In our simulation model with scouting only (A), foragers leave the hive on a search path (white lines) and continue until they encounter a resource (circles, colours indicate different resource quality). When foraging with recruitment (B) foragers continue to identify resources in scouting trips (white lines) and convey this information on the dance floor (brown disc) where foragers can sample dances reporting on scouting and recruiting trips and follow these directions (yellow lines). (C) Complementary cumulative frequencies of foraging distances reported from scouting and recruit trips. Note the difference in the shape of the distributions. The scout distribution is best fit by an exponential (grey fit line), the recruit distribution by a Rayleigh distribution (black line)"}
knitr::include_graphics("../Results/figures/manuscript_figures/fig2.png")
```

@@

```{r, fig.cap="Figure 3. **The rationale of the foraging model**. The distances of resources encountered by scouts are distributed exponentially (A). These dances are advertised on the dance floor (B). Dances for resources that are closer or higher in quality are repeated more often (C). As a consequence, dances for more profitable resource are over-represented and sampling foragers are biased to the more profitable resources (D). After successfully visiting advertised resources, recruits also dance for them leading to further amplification of this bias towards the most profitable resource in the vicinity of the hive (E). The distances of recruiting trips are than distributed through a Rayleigh distribution (F). Recruits also report the locations on the dance floor (B) and repeat their runs more often depending on the profitability of the location (C), leading to a distribution of durations of recruit dances (G). By taking together the dance distributions for scouts (D) and recruits (G) the distributions of all dances on the dance-floor can be found. The distances reported on the dance floor this are a mixture of the scout and recruiting trips and can be calculated from the distance distributions of the scouting and recruiting trips, taking the reporting bias into account (see Materials and Methods for detail)."}
knitr::include_graphics("../Results/figures/manuscript_figures/fig3.png")
```

To evaluate the use of waggle dance information and individual search in honeybee colonies foraging in 'natural' landscapes, we analysed a pre-existing dataset of 2827 waggle dance observations from 20 observation hives, recorded between April-September 2017, (previously described in [@Samuelson2021]). Hives that contributed to this dataset had been situated at different locations in South East England (see Materials and Methods, figure 4A) and visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data was recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually [@Couvillon2012;@Schurch2019;@Samuelson2021] to extract waggle run durations. Using this data, for each site we fitted both the collective and individual models and used model selection to determine which provided the better explanation of the data, and (if the collective model provided a better fit) to quantify the relative use of social information through estimating the parameter $p$. In each case, we calculated the goodness-of-fit using a Kolmogorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data [@Goldstein2004;@Clauset2009].

@@

```{r, fig.cap="Figure 4. **The honey bee foraging model fitted to data from 20 hives**. (A) Location of study hives in Southern England, shaded area in the main plot indicates Greater London. For 16 hives for the collective foraging model provided best explanation (black circles) or for 4 hives the individual search model provided the best explanation (red triangles) as indicated by lowest AIC score. (B). Distribution of goodness of fit confidence values for each model fit to waggle run durations from each site. The p-value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data, the red dashed line marks the significance threshold of 0.05. For values exceeding the threshold there is no statistically significant difference between the model and the data, indicating the model provides a good fit. For the hive in (C) the individual model (blue line) provided a better fit than the collective foraging model (red line). For the hive in (D) the collective foraging model (red line) provided a better fit than the individual model (blue line). The typical “hump” in the distribution in (D) which is indicative of contribution of recruitment dances (compare to Fig 3H). Panels show the compliment cumulative frequencies with binned frequency distributions as inset."}
knitr::include_graphics("../Results/figures/manuscript_figures/fig4.png")
```

@@

quantifying the use of waggle-dance recruitment within all colonies, as a proportion of all foraging trips, can be achieved by extracting the estimated proportion of recruit trips, $1-p$, for each site. Since our sites varied in land-use characteristics, and thus potentially forage distributions (although in this case, not by design for this study), we investigated whether these estimates might correlate with land-use. We first classified the different land-use types of the area surrounding each site [@Samuelson2018] to obtain a standardised land-use profile for the urban and agri-rural environments separately as many land-use types present in urban areas do not occur in agri-rural environments and vice versa (see Materials and Methods). We then performed a Partial Least Squares (PLS) analysis [@Carrascal2009] (see Materials and Methods) to determine the principal components that represent combinations of land-use types which explained the most variation in waggle dance use within agri-rural and urban environments. As for one of our sites neither model provided a plausible description, this site was removed from the PLS analysis.

@@

```{r, fig.cap="Figure 5. **Collective foraging correlates with land-use**. Estimated waggle dance use for each site against the first principal component derived from a Partial Least Squares analysis of land-use type. Beta regression shows the relationship (black line) between first principal component and waggle dance use, with 95% CI shown by the grey shaded area for agri-rural (A) and urban landscapes (C). The correlations between first principal component and each land-use type are shown for agri-rural (B) and urban landscapes (D). Correlations outside the shaded area significantly contribute to the first principal component. Colours correspond to the land use as shown in maps (circular insets) for selected sites. NAUMGS (resp. NAMGS) stands for non-agricultural unmanaged (resp. managed) green space"}
knitr::include_graphics("../Results/figures/manuscript_figures/fig5.png")
```

@@

load("../Results/area_coverage.RData")

@@

Goodness of fit is assessed using the two-sample Kolmogorov-Smirnov (KS) test [@Goldstein2004] and implemented in R using the ks.boot function of the package 'Matching' in R [@Sekhon2011].

@@

All of the data and code required to reproduce this article is contained as within the 'wagglefit' which can be found at (link to finalised Zenodo archive finalised)

https://github.com/joseph-palmer/wagglefit
